### 2025-06-06 paper/conclusion.tex
Update conclusion, introduction, results, and thesis abstract to reflect changes in analysis methodology from two-staged to ex-post.

Our methodological contributions extend beyond the specific applications studied. The two-staged analysis methodology provides systematic approaches for extracting insights from multi-agent learning dynamics despite catastrophic forgetting in neural networks. The discretization approach for continuous-time economic models, treating Lévy processes through Euler-Maruyama schemes and mapping continuous decisions to discrete action probabilities, provides a template for computational implementation while preserving essential incentive structures. We constructed an observed reward function that enables reinforcement learning in environments without direct reward signals and developed a specialized transformer loss function for Lévy process observations. Our implementation incorporates Graph Neural Networks with temporal and spatial attention mechanisms that capture both network topology and temporal dependencies in multi-agent interactions. Furthermore, we introduce off-equilibrium asymmetric analysis methods that move beyond the symmetric equilibria analysis in economic theory, allowing for heterogeneous agent behaviors and asymmetric strategy convergence patterns.

>>>>

Our methodological contributions extend beyond the specific applications studied. The ex-post analysis methodology provides systematic approaches for extracting insights from multi-agent learning dynamics despite catastrophic forgetting in neural networks. The discretization approach for continuous-time economic models, treating Lévy processes through Euler-Maruyama schemes and mapping continuous decisions to discrete action probabilities, provides a template for computational implementation while preserving essential incentive structures. We constructed an observed reward function that enables reinforcement learning in environments without direct reward signals and developed a specialized transformer loss function for Lévy process observations. Our implementation incorporates Graph Neural Networks with temporal and spatial attention mechanisms that capture both network topology and temporal dependencies in multi-agent interactions. Furthermore, we introduce off-equilibrium asymmetric analysis methods that move beyond the symmetric equilibria analysis in economic theory, allowing for heterogeneous agent behaviors and asymmetric strategy convergence patterns.

### 2025-06-06 paper/intro.tex
Update conclusion, introduction, results, and thesis abstract to reflect changes in analysis methodology from two-staged to ex-post.

Third, we introduce POLARIS (Partially Observable Learning with Active Reinforcement In Social environments), combining belief processing through Transformer models, variational inference learning with Graph Neural Networks (GNNs), and reinforcement learning optimization. Our implementation incorporates GNNs with temporal and spatial attention mechanisms that capture network topology and temporal dependencies, enabling agents to develop sophisticated strategies accounting for both environmental partial observability and strategic adaptation. POLARIS addresses the critical challenge of catastrophic forgetting through a two-staged analysis methodology that provides systematic approaches for extracting insights from multi-agent learning dynamics despite agents' inability to maintain stable policies across all states.

>>>>

Third, we introduce POLARIS (Partially Observable Learning with Active Reinforcement In Social environments), combining belief processing through Transformer models, variational inference learning with Graph Neural Networks (GNNs), and reinforcement learning optimization. Our implementation incorporates GNNs with temporal and spatial attention mechanisms that capture network topology and temporal dependencies, enabling agents to develop sophisticated strategies accounting for both environmental partial observability and strategic adaptation. POLARIS addresses the critical challenge of catastrophic forgetting through an ex-post analysis methodology that provides systematic approaches for extracting insights from multi-agent learning dynamics despite agents' inability to maintain stable policies across all states.

### 2025-06-06 paper/results.tex
Update conc, intro, abs.
Update conclusion, introduction, results, and thesis abstract to reflect changes in analysis methodology from two-staged to ex-post.

These fundamental constraints require us to reset agent parameters at the beginning of each episode rather than allowing continuous learning across state realizations. While this methodological necessity prevents direct measurement of the equilibrium strategies that economic theory and our POAMG framework predict, it does not invalidate our computational approach. Rather, it compels us to develop alternative analytical methodologies that can extract meaningful insights about multi-agent strategic behavior despite these inherent limitations of applying reinforcement learning to environments with conflicting learning objectives.

Our experimental methodology therefore focuses on within-episode learning dynamics and cross-episode pattern analysis. By examining how agents adapt their behavior within episodes and identifying systematic patterns across episodes, we can still characterize the essential features of multi-agent social learning, even when direct equilibrium convergence measurement is precluded by computational limitations.

\subsection{Two-Staged Analysis Methodology}
\label{sec:two-staged-analysis}

To address the catastrophic forgetting limitation while still extracting meaningful insights about multi-agent learning patterns, we adopt a consistent two-staged analysis approach across both experimental scenarios. This methodology enables robust characterization of learning dynamics despite the fundamental constraints of applying reinforcement learning to environments with fixed but episode-varying states.

@@

This methodology reveals whether observed disparities represent structural features of multi-agent learning or merely artifacts of specific agent characteristics like network position, as different agents assume extreme roles across different episodes.

@@

Following our two-staged analysis methodology, the POLARIS implementation tracks: (1) Convergence of individual agent allocation strategies to state-dependent optimal actions within episodes; (2) Belief fluctuations over time to assess the impact of neighbor action observation on learning; (3) Cumulative allocation disparities between highest and lowest allocators as a function of group size; and (4) Episode-wise analysis of allocation patterns across different true states (good vs. bad).

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\textwidth]{../code/polaris/docs/highest_lowest_cumulative_allocators_bad_state.png}
    \caption{Highest and lowest cumulative allocators in the bad state. (Average over 20 episodes)}
    \label{fig:highest_lowest_cumulative_allocators_bad_state}
\end{figure}
\paragraph{Dynamic Role Assignment and Allocation Disparities} Figures \ref{fig:highest_lowest_cumulative_allocators_good_state} and \ref{fig:highest_lowest_cumulative_allocators_bad_state} demonstrate a systematic widening of the gap between highest and lowest cumulative allocators as group size increases. This pattern reveals the emergence of dynamic role assignment in multi-agent learning, where agents naturally differentiate into information generators and information exploiters within each episode. In strategic experimentation, this manifests as some agents reducing individual experimentation when benefiting from others' information generation (free-riding effects), while enabling top experimenters to allocate more aggressively when supported by group learning (encouragement effects). Interestingly, despite the presence of free-riding behavior typically associated with inefficiency in economic literature, our results show that cumulative allocation in larger networks substantially exceeds autarky levels in both good and bad states, suggesting that the social learning benefits and encouragement effects more than compensate for the traditional inefficiencies of free-riding in these multi-agent learning environments.

@@

Following our two-staged analysis methodology described in Section \ref{sec:two-staged-analysis}, we track each agent's incorrect action probability as the probability assignment of the policy network to the incorrect action over time and estimate individual learning rates by fitting log-linear regressions to identify the fastest and slowest learners within each episode. We then aggregate the performance of these extreme learners across episodes to examine whether the observed learning disparities represent structural features of multi-agent learning rather than artifacts of specific agent characteristics.

>>>>

These fundamental constraints force us to reset agent parameters at the beginning of each episode rather than allowing continuous learning across state realizations—a methodological compromise that fundamentally undermines our ability to directly test the theoretical equilibrium conditions. This parameter reset prevents agents from developing the ex ante conditional strategies that both economic theory and our POAMG framework predict, instead forcing them to learn state-specific policies that cannot generalize across different environmental realizations.

Given these constraints, our experimental methodology necessarily shifts focus to within-episode learning dynamics and cross-episode pattern analysis. This indirect approach allows us to characterize certain features of multi-agent social learning, though it falls short of the direct equilibrium validation that would constitute ideal theoretical testing.

\subsection{Ex-Post Analysis Methodology}
\label{sec:ex-post-analysis}

Recognizing that catastrophic forgetting prevents direct measurement of equilibrium strategies, we implement an ex-post analysis methodology as a workaround to extract off-equilibrium learning patterns. While this approach cannot resolve the fundamental limitations discussed above, it provides a systematic framework for analyzing the learning patterns that we can observe within the constraints of our computational approach.

@@

This methodology provides a limited window into multi-agent learning patterns, allowing us to examine whether observed disparities represent structural features that persist across episodes rather than artifacts of specific agent characteristics.

@@

Following our ex-post analysis methodology, the POLARIS implementation tracks: (1) Convergence of individual agent allocation strategies to state-dependent optimal actions within episodes; (2) Belief fluctuations over time to assess the impact of neighbor action observation on learning; (3) Cumulative allocation disparities between highest and lowest allocators as a function of group size; and (4) Episode-wise analysis of allocation patterns across different true states (good vs. bad).

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\textwidth]{../code/polaris/docs/highest_lowest_cumulative_allocators_bad_state.png}
    \caption{Highest and lowest cumulative allocators in the bad state. (Average over 20 episodes)}
    \label{fig:highest_lowest_cumulative_allocators_bad_state}
\end{figure}

\paragraph{Dynamic Role Assignment and Allocation Disparities} Figures \ref{fig:highest_lowest_cumulative_allocators_good_state} and \ref{fig:highest_lowest_cumulative_allocators_bad_state} demonstrate a systematic widening of the gap between highest and lowest cumulative allocators as group size increases. This pattern reveals the emergence of dynamic role assignment in multi-agent learning, where agents naturally differentiate into information generators and information exploiters within each episode. In strategic experimentation, this manifests as some agents reducing individual experimentation when benefiting from others' information generation (free-riding effects), while enabling top experimenters to allocate more aggressively when supported by group learning (encouragement effects). Interestingly, despite the presence of free-riding behavior typically associated with inefficiency in economic literature, our results show that cumulative allocation in larger networks substantially exceeds autarky levels in both good and bad states, suggesting that the social learning benefits and encouragement effects more than compensate for the traditional inefficiencies of free-riding in these multi-agent learning environments.

@@

Following our ex-post analysis methodology described in Section \ref{sec:ex-post-analysis}, we track each agent's incorrect action probability as the probability assignment of the policy network to the incorrect action over time and estimate individual learning rates by fitting log-linear regressions to identify the fastest and slowest learners within each episode. We then aggregate the performance of these extreme learners across episodes to examine whether the observed learning disparities represent structural features of multi-agent learning rather than artifacts of specific agent characteristics.

### 2025-06-06 paper/thesis.tex
Update appendices
Update results.
Update conclusion, introduction, results, and thesis abstract to reflect changes in analysis methodology from two-staged to ex-post.

% Abstract
\begin{abstract}
\begin{adjustwidth}{2cm}{2cm}
This thesis bridges economic social learning theory and multi-agent reinforcement learning by developing Partially Observable Active Markov Games (POAMGs) and the POLARIS algorithm. Our framework addresses fundamental challenges in modeling adaptive behavior under partial observability, where agents must simultaneously learn about uncertain environments and evolving strategies of others. We extend Active Markov Games to partially observable settings, derive theoretical convergence results, and develop a practical algorithm combining Transformer-based belief processing, Graph Neural Networks with attention mechanisms, and reinforcement learning optimization. Applied to strategic experimentation and learning without experimentation scenarios, our approach reveals dynamic role assignment as a robust organizing principle where agents naturally differentiate into complementary roles enhancing collective information processing. We find that free-riding behavior does not lead to uniform inefficiencies—for some agents, performance in larger networks exceeds autarky levels. Our two-staged analysis methodology addresses computational challenges like catastrophic forgetting while providing systematic approaches for extracting insights from multi-agent learning dynamics. The framework contributes theoretical foundations to multi-agent reinforcement learning and computational tools to economic theory, demonstrating how interdisciplinary approaches can illuminate complex social learning phenomena neither field could address alone.
\end{adjustwidth}
\end{abstract}

@@

% Page numbering starts here with Arabic numerals
\chapter{Introduction}
\pagenumbering{arabic}
\setcounter{page}{1}
\input{intro}
\chapter{Related Work}
\input{background}
\chapter{Reinforcement Learning Framework}
\input{theory}
\chapter{Evaluation}
\input{results}
\chapter{Conclusion}
\input{conclusion}

% Appendix
\newpage
\appendix
\addcontentsline{toc}{chapter}{Appendices}
\input{appendix}

>>>>

% Abstract
\begin{abstract}
\begin{adjustwidth}{2cm}{2cm}
This thesis bridges economic social learning theory and multi-agent reinforcement learning by developing Partially Observable Active Markov Games (POAMGs) and the POLARIS algorithm. Our framework addresses fundamental challenges in modeling adaptive behavior under partial observability, where agents must simultaneously learn about uncertain environments and evolving strategies of others. We extend Active Markov Games to partially observable settings, derive theoretical convergence results, and develop a practical algorithm combining Transformer-based belief processing, Graph Neural Networks with attention mechanisms, and reinforcement learning optimization. Applied to strategic experimentation and learning without experimentation scenarios, our approach reveals dynamic role assignment as a robust organizing principle where agents naturally differentiate into complementary roles enhancing collective information processing. We find that free-riding behavior does not lead to uniform inefficiencies—for some agents, performance in larger networks exceeds autarky levels. Our ex-post analysis methodology addresses computational challenges like catastrophic forgetting while providing systematic approaches for extracting insights from multi-agent learning dynamics. The framework contributes theoretical foundations to multi-agent reinforcement learning and computational tools to economic theory, demonstrating how interdisciplinary approaches can illuminate complex social learning phenomena neither field could address alone.
\end{adjustwidth}
\end{abstract}

@@

% Page numbering starts here with Arabic numerals
\chapter{Introduction}
\pagenumbering{arabic}
\setcounter{page}{1}
\input{intro}
\chapter{Related Literature}
\input{background}
\chapter{Reinforcement Learning Framework}
\input{theory}
\chapter{Evaluation}
\input{results}
\chapter{Conclusion}
\input{conclusion}

% Appendix
\newpage
\appendix
\addcontentsline{toc}{chapter}{Appendices}

% Appendices title page
\begin{center}
\vspace*{5cm}
{\Huge \textbf{Appendices}}
\end{center}
\newpage

\input{appendix}
